Rectangulation polytopes

### 2024-03-20 draft/rectangulotopes.tex
additional statements

\documentclass[12pt]{amsart}

@@

@@

\begin{abstract}
  Rectangulations are decompositions of a square into finitely many axis-aligned rectangles.
  We give simple realizations of $(n-1)$-dimensional polytopes associated with rectangulations composed of $n$ rectangles. They correspond to quotientopes of natural congruence relations on the weak Bruhat order on permutations in $\f{S}_n$, and their skeleta are flip graphs on rectangulations. We give  elementary formulas for computing the coordinates of each vertex of the polytope, in the spirit of Loday's realization of the associahedra.
\end{abstract}

@@

@@

\begin{itemize}
\item Terminology: Rectangles, vertices, segments
\item Wall slides
\item Two equivalence classes: weak equivalence preserving rectangle-segment adjacencies, and strong equivalence preserving rectangle-rectangle adjacencies.
\item Weak equivalence classes have a \emph{diagonal} representative.
\end{itemize}

@@

\begin{theorem}
  Let $\WRP (n)$ be the convex hull of the points
  $\sum_{i\in [n]} w_i\cdot \mathbf{e}_i$,
  where
  \[
  \begin{split}
    w_i := \quad & h(T(R), i)\cdot v(T(R),i) \\
           \quad & - h(S(R), i)\cdot v(S(R),i),
  \end{split}
  \]
  for all weak rectangulations $R$ on $n$ rectangles, where $h(T,i)$ and $v(T,i)$ denote respectively the number of leaves in the horizontal and vertical subtrees of $i$ in the tree $T$.
  Then $\WRP (n)$ is a realization of the $(n-1)$-dimensional weak rectangulation polytope.
\end{theorem}

@@

Our second main result on the realizations of the strong rectangulation polytopes does not require any additional definition. It is, however, slightly more involved.

\begin{theorem}
  Let $\SRP (n)$ be the convex hull of the points
  \[
  \sum_{i,j\in [n]} w_{i,j}\cdot (\mathbf{e}_i - \mathbf{e}_j),
  \]
  with
  \[
  \begin{split}
  w_{i,j} := \quad & h(T(R), i) \cdot cv (T(R), S(R), i, j)\cdot h(S(R), j) \\
            \quad & - v(S(R), j) \cdot ch (T(R), S(R), i, j)\cdot v(T(R), i)
  \end{split}
  \]
  \jean{Write the correct formula here, and sum only on the suitable pairs $i,j$.}
  for all strong rectangulations $R$ on $n$ rectangles, where:
  \begin{itemize}
  \item $h(T,i)$ and $v(T,i)$ denote respectively the number of leaves in the horizontal and vertical subtrees of $i$ in the tree $T$,
  \item $ch (T,S,i,j)$ and $cv (T,S,i,j)$ denote respectively the number of \emph{common leaves} to the horizontal (vertical) subtree of $i$ in $T$ and the horizontal (vertical) subtree of $j$ in $S$
  \end{itemize}
Then $\SRP (n)$ is a realization of the $(n-1)$-dimensional strong rectangulation polytope.  
\end{theorem}

@@

\jean{State the main results that we will use. I just give sloppy statements for now.}

@@

Define the weak poset $P_w(R)$ of a rectangulation $R$ as the intersection of two partial orders defined by the twin binary trees

\begin{theorem}
  The following subsets of $\f{S}_n$ 
  \[
  \{\mathcal{L}(P_w(R)) : R \text{ is a weak rectangulation} \}
  \]
  are equivalence classes of a congruence relation on the weak Bruhat order on $\f{S}_n$.
  The minimal elements of each equivalence classes are twisted Baxter permutations, and the maximal
  elements are co-twisted Baxter permutations.
\end{theorem}

@@

We call this congruence the \emph{Baxter congruence}.
The weak rectangulation polytope $\WRP(n)$ is the quotientope of the Baxter congruence of the weak Bruhat order on permutations.

@@

Define the strong poset $P_s(R)$ of a rectangulation $R$ using the intertwined binary trees.

\begin{theorem}
  The following subsets of $\f{S}_n$ 
  \[
  \{\mathcal{L}(P_s(R)) : R \text{ is a strong rectangulation} \}
  \]
  are equivalence classes of a congruence relation on the weak Bruhat order on $\f{S}_n$.
  The minimal elements of each equivalence classes are 2-clumped permutations, and the maximal
  elements are co-2-clumped permutations.
\end{theorem}

The strong rectangulation polytope $\SRP(n)$ is the quotientope of the congruence of the weak Bruhat order on permutations that is defined by the 2-clumped permutations.

\jean{Does the congruence have a little name?}

@@

The congruence is defined by forbidding the arcs that cross the horizontal line more than once.
The quotientope is therefore the Minkowski sum of the shard polytopes corresponding to four families of arcs:
\begin{itemize}
\item the Loday and anti-Loday arcs, like in the weak case,
\item the Yin and yang arcs.
\end{itemize}

@@

Proof of our second main result:
\begin{itemize}
\item Identify what are the vertices of each yin summands that contribute to a given rectangulation,
\item group the contributions for each pair of rectangles $i,j$,
\item prove that the count is the triple product of numbers of leaves of the trees,
\item argue about what pairs $i,j$ are to be considered,
\item argue symmetrically about yang arcs and conclude.
\end{itemize}

>>>>

\documentclass{amsart}

@@

% shard polytopes
\newcommand{\SP}{\polytope{SP}}

% Loday and anti-Loday accents (?)
\newcommand{\loday}[1]{\overset{\frown}{#1}}
\newcommand{\antiloday}[1]{\overset{\smile}{#1}}
% yin and yang accents (?)
\newcommand{\yang}[1]{\overset{\backsim}{#1}}
\newcommand{\yin}[1]{\overset{\sim}{#1}}

@@

\begin{abstract}
  Rectangulations are decompositions of a square into finitely many axis-aligned rectangles.
  We describe realizations of $(n-1)$-dimensional polytopes associated with rectangulations composed of $n$ rectangles. They are defined as quotientopes of natural congruence relations on the weak Bruhat order on permutations in $\f{S}_n$, and their skeleta are flip graphs on rectangulations. We give  elementary formulas for computing the coordinates of each vertex of the polytope, in the spirit of Loday's realization of the associahedra.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage

@@

\begin{itemize}
\item Terminology: rectangles, vertices, segments, generic rectangulation
\item Wall slides
\item Two equivalence classes: \defn{weak equivalence} preserving rectangle-segment adjacencies, and \defn{strong equivalence} preserving rectangle-rectangle adjacencies.
\item Weak equivalence classes have a \defn{diagonal} representative.
\end{itemize}

@@

\begin{theorem}
  Let $\WRP (n)$ be the convex hull of the points
  $\sum_{i\in [n]} (\loday{w}_i - \antiloday{w}_i)\cdot \mathbf{e}_i$,
  where
  \[
  \begin{split}
    \loday{w}_i := & h(T(R), i)\cdot v(T(R),i) \\
    \antiloday{w}_i := & h(S(R), i)\cdot v(S(R),i),
  \end{split}
  \]
  for all weak rectangulations $R$ on $n$ rectangles, where $h(T,i)$ and $v(T,i)$ denote respectively the number of leaves in the horizontal and vertical subtrees of $i$ in the tree $T$.
  Then $\WRP (n)$ is a realization of the $(n-1)$-dimensional weak rectangulation polytope.
\end{theorem}

@@

Two leaves of the trees $T(R)$ and $S(R)$ are said to be \defn{common leaves} if they are both either horizontal or vertical, and the edges to their parents lie on the same segment of $R$. We also denote by $i|j$ and $\frac{\ j\ }i$ the situations where the rectangle labeled $i$ touches the rectangle labeled $j$ from the left and from below in $R$, respectively.

\begin{theorem}
  Let $\SRP (n)$ be the convex hull of the points
  \[
  \sum_{i,j\in [n], i< j} (\yin{w}_{i,j} - \yang{w}_{i,j})\cdot (\mathbf{e}_i - \mathbf{e}_j),
  \]
  with
  \[
  \yin{w}_{i,j} := 
  \begin{cases}
    h(T(R), i) \cdot cv (T(R), S(R), i, j)\cdot h(S(R), j) & \text{ if neither } i|j \text{ nor } \frac{\ j\ }i \\
    0 & \text{ otherwise }
  \end{cases}
  \]
  and
    \[
  \yang{w}_{i,j} := 
  \begin{cases}
    v(S(R), j) \cdot ch (T(R), S(R), i, j)\cdot v(T(R), i) & \text{ if neither } j|i \text{ nor } \frac{\ i\ }j \\
    0 & \text{ otherwise }
  \end{cases}
  \]

  \jean{Double check the formula.}
  for all strong rectangulations $R$ on $n$ rectangles, where:
  \begin{itemize}
  \item $h(T,i)$ and $v(T,i)$ denote respectively the number of leaves in the horizontal and vertical subtrees of $i$ in the tree $T$,
  \item $ch (T,S,i,j)$ and $cv (T,S,i,j)$ denote respectively the number of common leaves of the horizontal (vertical) subtree of $i$ in $T$ and the horizontal (vertical) subtree of $j$ in $S$
  \end{itemize}
Then $\SRP (n)$ is a realization of the $(n-1)$-dimensional strong rectangulation polytope.  
\end{theorem}

\Jean{Is there a more concise way to write this down/compute? Here we do not take advantage of the fact that at most one of the two factors $yin{w}_{i,j}$ and $\yang{w}_{i,j}$ is nonzero.}

@@

@@

Define the weak poset $P_w(R)$ of a rectangulation $R$ as the union of two partial orders defined by the twin binary trees

\begin{theorem}
  The subsets $\{\mathcal{L}(P_w(R)) : R \text{ is a rectangulation} \}$ of $\f{S}_n$ 
  are equivalence classes of a congruence relation on the weak Bruhat order on $\f{S}_n$.
  The minimal elements of each equivalence classes are twisted Baxter permutations, and the maximal
  elements are co-twisted Baxter permutations.
\end{theorem}

@@

We call this congruence the \emph{Baxter congruence}.
The weak rectangulation polytope $\WRP(n)$ is the quotientope of the Baxter congruence of the weak Bruhat order on $\f{S}_n$.

@@

Define the strong poset $P_s(R)$ of a rectangulation $R$. Can we use the intertwined binary trees?

\begin{theorem}
  The subsets $\{\mathcal{L}(P_s(R)) : R \text{ is a rectangulation} \}$ of $\f{S}_n$ 
  are equivalence classes of a congruence relation on the weak Bruhat order on $\f{S}_n$.
  The minimal elements of each equivalence classes are 2-clumped permutations, and the maximal
  elements are co-2-clumped permutations.
\end{theorem}

The strong rectangulation polytope $\SRP(n)$ is the quotientope of the congruence of the weak Bruhat order on $\f{S}_n$ that is defined by the 2-clumped permutations.

\jean{Does the congruence have a name?}

@@

The congruence is defined by forbidding the arcs that cross the horizontal line more than once.
The quotientope is therefore the Minkowski sum of the shard polytopes corresponding to four families of arcs:
\begin{itemize}
\item the Loday and anti-Loday arcs, like in the weak case,
\item the yin and yang arcs.
\end{itemize}

@@

Proof of our second main result:
\begin{itemize}
\item Identify what are the yin summands contributing to a given pair $\mathbf{e}_i - \mathbf{e}_j$,
\item prove that the count is the triple product of numbers of leaves of the trees,
\item argue about what pairs $i,j$ are to be considered,
\item argue symmetrically about yang arcs and conclude.
\end{itemize}

\begin{lemma}
  \label{lem:yinminmax}
  Let $\pi\in\f{S}_n$ and consider a yin arc $A$ composed of two intervals $I$ and $J$.
  Let $i := \arg\max_{k\in I} \pi^{-1}(k)$ and $j := \arg\min_{k\in J} \pi^{-1}(k)$.
  The vertex of $\SP(A)$ that is extreme with respect to the direction $\pi^{-1}$
  is $\mathbf{e}_i-\mathbf{e}_j$ if $\pi^{-1}(i)>\pi^{-1}(j)$, and $0$ otherwise.
\end{lemma}

\begin{lemma}
  Let $R$ be a rectangulation on $n$ rectangles and $p(R)$ the vertex of $\SRP(n)$ associated with $R$.
  Given a pair $i,j\in [n]$ with $i<j$, the number of yin arcs $A$ such that $\mathbf{e}_i-\mathbf{e}_j$
  is the extremal vertex of $\SP(A)$ contributing to $p(R)$ is
  \[
  \yin{w}_{i,j} = 
  \begin{cases}
    h(T(R), i) \cdot cv (T(R), S(R), i, j)\cdot h(S(R), j) & \text{ if neither } i|j \text{ nor } \frac{\ j\ }i \\
    0 & \text{ otherwise }
  \end{cases}
  \]
\end{lemma}
\begin{proof}
  Let $\pi$ be the 2-clumped permutation corresponding to $R$, hence the minimal element of $\mathcal{L}(P_s(R))$ in the weak Bruhat order.
  From Lemma~\ref{lem:yinminmax}, the set of yin arcs $A$ such that $\SP(A)$ contributes to $\mathbf{e}_i-\mathbf{e}_j$
  are defined by pairs of contiguous intervals $I,J$ of $[n]$ with $i\in I$, $j\in J$, such that:
  \begin{itemize}
  \item $i = \arg\max_{k\in I} \pi^{-1}(k)$,
  \item $j = \arg\min_{k\in J} \pi^{-1}(k)$, and
  \item $i,j$ is an inversion in $\pi$, hence $\pi^{-1}(i)>\pi^{-1}(j)$.
  \end{itemize}
  We observe that provided the third condition holds, the number of choices of the three endpoints defining such a pair of intervals $I$ and $J$ is the product of the
  number of leaves of the subtree of $T(R)$ rooted at $i$, the number of common leaves to the vertical subtrees of $T(R)$ and $S(R)$ rooted
  at $i$ and $j$, respectively, and the number of leaves of the subtree of $S(R)$ rooted at $j$.

  It remains to observe that the only case where this product is nonzero and $i,j$ is not an inversion in $\pi$ is when either $i|j$ or $\frac{\ j\ }i$.
\end{proof}
